---
title: "Tornadoes and floods dealt the largest damage to the economy and health respectively
- USA, 1996 - 2011"
output: html_document
keep_md: yes
---

## Synopsis
This document briefly analyzes the impact of different weather events that occurred
in the United States between 1996 and 2011. More specifically, the economic
damage as represented by the damage to crops and property and the damage to population 
health as represented by the number of injuries and fatalities are considered. 
Since the purpose of the analysis is to help understand differences in
weather events, the events are compared based on the results of the damage analysis. 
By far the largest harm to population health was done by tornadoes. A second 
group of events that are relatively harmful to health is comprised of Excessive Heat,
Thunderstorm Wind, Flood, and Lightning. Regarding total economic damage the top causes are floods, hurricanes and storm surges. The same is true with respect to property damage. The top cause of crop damage is drought.

## Data processing
```{r}
# Functions and libraries
library(ggplot2)
library(plyr)
library(reshape2)
library(knitr)
```

Downloading and reading in the data (source: NOAA):

```{r, cache=TRUE}
if (!file.exists("stormdata.csv.bz2")){
      download.file(url = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormdata.csv.bz2")
      }

# unzip
if (!file.exists("stormdata.csv")){
      library(R.utils)
      bunzip2("stormdata.csv.bz2", destname = "stormdata.csv", remove=F)
      }
stormdata <- read.csv("stormdata.csv")
```

As stated by the NOAA, weather events of all types were recorded only since 1996,
see http://www.ncdc.noaa.gov/stormevents/details.jsp. During most of the years
in the data set only tornadoes were recorded. Thus, all years prior to 1996 
will be dropped in order to avoid the biasing of subsequent estimates.

```{r, cache=TRUE}
# Format date variable
stormdata$BGN_DATE <- as.character(stormdata$BGN_DATE)
# Function to select only the date part
f <- function(x) strsplit(x, split = " ")[[1]][1]
stormdata$BGN_DATE <- sapply(stormdata$BGN_DATE, FUN = f); rm(f)
stormdata$BGN_DATE <- as.Date(stormdata$BGN_DATE, format = "%m/%d/%Y")
stormdata$year <- format(stormdata$BGN_DATE, format = "%Y")
# Keep only if event from 1996 or later
stormdata <- stormdata[which(stormdata$year >= 1996), ]
```

Since only events that caused injuries, fatalities or damage are relevant here, other events can be dropped from the data set.

```{r}
nofatalities <- which(stormdata$FATALITIES == 0)
noinjuries <- which(stormdata$INJURIES == 0)
nocropdmg <- which(stormdata$CROPDMG == 0)
nopropdmg <- which(stormdata$PROPDMG == 0)

drop <- Reduce(intersect, list(nofatalities, noinjuries, nocropdmg, nopropdmg))
rm(nofatalities, noinjuries, nocropdmg, nopropdmg)
```

Of `r nrow(stormdata)` records `r length(drop)` can be dropped.

```{r}
stormdata <- stormdata[-drop, ]; rm(drop)
```

For economic damage the PROPDMGEXP (exponent) variable needs to be taken into account. 
The same applies for crop damage and the CROPDMGEXP variable. The damage has to be multiplied accordingly (e.g. multiply by 1000 if EXP = K).

```{r}
# Calculate fatalities and injuries per event type
fatalities <- tapply(X = stormdata$FATALITIES, INDEX = stormdata$EVTYPE, FUN = sum)
fatality_evtypes <- names(fatalities)
injuries <- tapply(X = stormdata$INJURIES, INDEX = stormdata$EVTYPE, FUN = sum)

# Create a multiplier variable for property damage, because it is a character
# in the original data
stormdata$propfactor <- 1
expcodes <- c("h", "H", "k", "K", "m", "M", "b", "B")
exponents <- c(100, 100, 1000, 1000, 1000000, 1000000, 1000000000, 1000000000)
for (i in seq_along(expcodes)){
      index <- which(stormdata$PROPDMGEXP == expcodes[i])
      stormdata$propfactor[index] <- exponents[i]
      }

# The same for crop damage
stormdata$cropfactor <- 1
expcodes <- c("h", "H", "k", "K", "m", "M", "b", "B")
exponents <- c(100, 100, 1000, 1000, 1000000, 1000000, 1000000000, 1000000000)
for (i in seq_along(expcodes)){
      index <- which(stormdata$CROPDMGEXP == expcodes[i])
      stormdata$cropfactor[index] <- exponents[i]
      }
rm(expcodes, exponents, i, index)

# Damage in billions of dollars
stormdata$propertydamage <- (stormdata$PROPDMG * stormdata$propfactor) / 1e+09
stormdata$cropdamage <- (stormdata$CROPDMG * stormdata$cropfactor) / 1e+09
propertydmg <- tapply(X = stormdata$propertydamage, INDEX = stormdata$EVTYPE, FUN = sum)
cropdmg <- tapply(X = stormdata$cropdamage, INDEX = stormdata$EVTYPE, FUN = sum)

# Calculate totals in order to display percentages of totals later
injuriessum <- sum(stormdata$INJURIES)
fatalitiessum <- sum(stormdata$FATALITIES)
cropdmgsum <- sum(stormdata$cropdamage)
propertydmgsum <- sum(stormdata$propertydamage)
```

The event types should belong to 48 standard types, see the National Weather Service Storm Data Documentation, page 6. There are much more in the data:

```{r}
stormdata$EVTYPE <- as.character(stormdata$EVTYPE)
length(unique(stormdata$EVTYPE))
```

In the subset of events after 1996 there are `r length(unique(stormdata$EVTYPE))` different types. Without domain knowledge it would be problematic to categorize all the rare events as one of the standard types, so this attempt was not made. However, among the most common events are "THUNDERSTORM WIND" and "TSTM WIND" which seem to be the same which is why all event types including the term "TSTM WIND" will be categorized as "THUNDERSTORM WIND". The same applies for "WILDFIRE" / "WILD/FOREST FIRE" and "RIP CURRENT" / "RIP CURRENTS".

```{r}
tstm_index <- grep(pattern = "TSTM WIND", x = stormdata$EVTYPE)
stormdata$EVTYPE[tstm_index] <- "THUNDERSTORM WIND"
rm(tstm_index)

stormdata$EVTYPE[stormdata$EVTYPE == "RIP CURRENTS"] <- "RIP CURRENT"

stormdata$EVTYPE[stormdata$EVTYPE == "WILD/FOREST FIRE"] <- "WILDFIRE"
```


## Results
After this reformatting the top 20 event types ordered by number of occurrences from 1996 until 2011 are the following:

```{r, results='asis'}
tabledata <- head(sort(table(stormdata$EVTYPE), decreasing = T), 20)
tabledata <- melt(tabledata)
colnames(tabledata) <- c("Event type", "Sum of occurrences")
kable(tabledata)
```

These top 20 events represent `r sum(tabledata[,"Sum of occurrences"])` of `r sum(table(stormdata$EVTYPE))` all recorded events in the data.

### Fatalities and injuries
In order to assess the impact of the different weather events fatalities and injuries are considered first. 

```{r}
# NOTE: This section was not included in the chapter on data processing 
#       because it is mainly about selecting from the processed data
# Select top causes
# At least n injuries + fatalities
suminjfat <- injuries + fatalities
suminjfat <- sort(suminjfat, decreasing = T)
suminjfat <- head(suminjfat, 20)
# Which are the top 20 events by sum of injuries and fatalities?
selected_events <- names(suminjfat)

# Select from injuries and fatalities data
injuries2 <- injuries[names(injuries) %in% selected_events] 
fatalities2 <- fatalities[names(fatalities) %in% selected_events] 

# Plot fatality and injury values depending on event type
m_injuries2 <- melt(injuries2, value.name="Injuries")
m_fatalities2 <- melt(fatalities2, value.name="Fatalities")
plotdata <- merge(m_fatalities2, m_injuries2, by = "Var1")
colnames(plotdata)[1] <- "Event Type"
plotdata <- melt(plotdata, id.vars = c("Event Type"), 
                 value.name = "Amount", 
                 variable.name = "Damagetype")

ybreaks <- 150 * 2^(0:10)
ggplot(plotdata, 
       aes(x=reorder(plotdata$`Event Type`, plotdata$Amount, FUN=sum, order=T),
           y=Amount)) + 
      geom_point(aes(colour=Damagetype), stat="identity", size=5) +
      scale_y_sqrt(breaks = ybreaks) + xlab("Event Type") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      ggtitle("Fatalities and injuries depending on the event type (1996 - 2011)")
```

Tornadoes caused by far the most injuries (please note the logarithmic scale). A second group of events consisting of excessive heat, flood, thunderstorm wind and lightning also caused a considerable amount of injuries. Excessive heat caused the most fatalities, followed by tornadoes. 

In general, all of the top 16 events cause more injuries than fatalities. One notable exception are flash floods which cause nearly as many fatalities as injuries so that the number of fatalities caused by flash floods is the third highest although regarding the sum of fatalities and injuries flash floods is at rank six.

```{r, results='asis'}
# Table
m_suminjfat <- melt(suminjfat)
# Calculate percentages
m_fatalities2$fatalitypercentage <- (m_fatalities2$Fatalities / fatalitiessum) * 100
m_injuries2$injurypercentage <- (m_injuries2$Injuries / injuriessum) * 100

tabledata <- join_all(list(m_fatalities2, m_injuries2, m_suminjfat), 
                      by = 'Var1', type = 'full')
colnames(tabledata) <- c("Event", "Fatalities", "...percent of total", 
                         "Injuries", "...percent of total", "Sum")
kable(tabledata[order(tabledata$Sum, decreasing = T), ], row.names = F, digits = 2)
```

### Economic damage
Economic damage is defined here as the sum of property damage and crop damage. As stated in the NOAA's documentation many of these numbers are estimates. 

```{r, fig.cap="Figure 2: Economic damage"}
# NOTE: This section was not included in the chapter on data processing 
#       because it is mainly about selecting from the processed data
## Select economically most harmful events
sumdmg <- propertydmg + cropdmg
sumdmg <- sort(sumdmg, decreasing = T)
sumdmg <- head(sumdmg, 20)
# Which are the top 20 events by sum of damage?
selected_events <- names(sumdmg)

# Select from cropdmg and propertydmg data
cropdmg2 <- cropdmg[names(cropdmg) %in% selected_events] 
propertydmg2 <- propertydmg[names(propertydmg) %in% selected_events] 

# Plot fatality and injury values depending on event type
m_cropdmg2 <- melt(cropdmg2, value.name="cropdmg")
m_propertydmg2 <- melt(propertydmg2, value.name="propertydmg")
plotdata <- merge(m_propertydmg2, m_cropdmg2, by = "Var1")
colnames(plotdata) <- c("Event Type", "Property damage", "Crop damage")
plotdata <- melt(plotdata, id.vars = c("Event Type"), value.name = "Dollars", 
                 variable.name = "Damagetype")

ybreaks <- 2 * 2^(0:10)
ggplot(plotdata, 
       aes(x=reorder(plotdata$`Event Type`, plotdata$Dollars, FUN=sum, order=T), 
           y=Dollars)) + 
      geom_point(aes(colour=Damagetype), stat="identity", size=5) +
      scale_y_sqrt(breaks = ybreaks) +
      xlab("Event Type") + ylab("Billions of Dollars") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      ggtitle("Economic damage depending on the event type (1996 - 2011)")
```

With respect to economic damage tornadoes are less important. Floods 
caused roughly twice the damage of the second most dangerous event, hurricanes (again, please note the log scale). The vast majority of the top events inflicts a much larger damage to property than to crops. The exception here are droughts which are by far the top cause of crop damage and which caused about 13\$bn of crop damage and just 1\$bn of property damage.

```{r, results='asis'}
# Table
m_sumdmg <- melt(sumdmg)
# Calculate percentages
m_propertydmg2$fatalitypercentage <- (m_propertydmg2$propertydmg / propertydmgsum) * 100
m_cropdmg2$injurypercentage <- (m_cropdmg2$cropdmg / cropdmgsum) * 100

tabledata <- join_all(list(m_propertydmg2, m_cropdmg2, m_sumdmg), 
                      by = 'Var1', type = 'full')
colnames(tabledata) <- c("Event", "Property damage (bn USD)", "...as percent of total",
                         "Crop damage (bn USD)", "...as percent of total", "Sum")
kable(tabledata[order(tabledata$Sum, decreasing = T), ], digits = 2, row.names = F)
```